---
title: "Preliminars"
output:
  html_notebook: default
  html_document: default
  pdf_document: default
---

Shiny is an R package that helps building interactive web applications (apps) straight from R. Make sure to have installed [R](https://www.r-project.org/) and [RStudio](https://www.rstudio.com/products/rstudio/download/#download).  We will need a set of packages in the workshop, you can install them opening an R session, connect to internet, and run

```{r,message=FALSE, warning=FALSE}
install.packages("shiny", "ggplot2", "plotly","DT","RColorBrewer","rsconnect")
```

A shiny app is a web page (UI) that us connected to a live R session (Server). One of the advantages is that there exists an easy way to share your apps. The easiest way is to host them in [shinyapps.io](shinyapps.io). If you want to be able to share your apps follow the next steps prior to the workshop:

1. Create a free account at [shinyapps.io](shinyapps.io)

2. From your profile you will need
  + name
  + token
  + secret
  
3. You can set your account information from RStudio running the next code
```{r,message=FALSE, warning=FALSE}
rsconnect::setAccountInfo(name='your name',
                          token='your token',
                          secret='your secret')
```
 Following these steps will allow you to share your app with others (e.g. [Saudi Arabian health profile](https://population-health.shinyapps.io/saudi-arabia-health-profile/)).
